For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/paper.tex
stuff

We will use Agda-like type theoretic syntax to work in the internal
language of models of ETT:
\begin{itemize}
\item Universes are written $\Set_i$. We usually omit level indices in this paper.
\item $\Pi$ types are notated as $(x : A)\ra B$, or as $A \ra B$ when
  non-dependent. We sometimes omit function arguments, by implicitly
  generalising over variables.
\item $\Sigma$-types, notated either as $(x : A)\times B$, or as
  $\sum\limits_{x} B$ when we want to leave the type of the first
  projection implicit. Projections are either named or given by
  $\proj_1$ and $\proj_2$. We use $A \times B$ for non-dependent
  pairs.
\item We will define indexed W-types \cite{indexedcont} by listing
  their constructors and using pattern matching to eliminate from
  them.
\item The equality (identity) type is written $t = u$, it has a
  constructor $\refl : t = t$, and equality reflection, hence we use
  the same $=$ sign for definitional equality. We occasionally
  indicate by $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the
  equalities $e_1$,\dots,$e_n$. We also have uniqueness of identity
  proofs (UIP), expressing $(e : t = t)\ra e = \refl$. Note that
  function extensionality, expressing
  $((x : A) \ra f\,x = g\,x) \ra f = g$ is derivable.
\end{itemize}

@@

    & (t\oldapp u)^\A\,\gamma := t^\A\,\gamma\,(u^\A\,\gamma) && &&  (t\oldapp u)^\D\,\gamma^D := t^\D\,\gamma^D\,(u^\D\,\gamma^D) && &&  (t\oldapp u)^\S\,\gamma^S := u^\S\,\gamma^S \# t^\S\,\gamma^\S\,(u^\A\,\gamma) \\

@@

\begin{lemma}
  In a model of ETT supporting the theory of ITT signatures,
  signatures and syntactic signatures are isomorphic.
\end{lemma}
\begin{proof}
  We can turn a $({sig},{sigc}) : \Sign$ into $\Con^\I$ by ${sig}\,\I$
  and an $\Omega:\Con^\I$ into a $\Sign$ by
  $(\lambda M.\ll\Omega\rr_M, \lambda M\,N\,f.f\,(\ll\Omega\rr_M) =
  (f\circ\ll\rr_M)\,\Omega = \ll\Omega\rr_N)$ where the equality in
  the second component comes from uniqueness of the recursor (we have
  to define composition of morphisms $\circ$ for this). The
  compositions of these two maps are the identities: $({sig},{sigc})$
  is mapped to
  $(\lambda M.\ll{sig}\,I\rr_M,\dots) = (\lambda
  M.\ll\blank\rr_M\,({sig}\,I),\dots)$ which is equal to
  $(\lambda M.{sig}\,M,\dots)$ because of ${sigc}$; $\Omega$ is mapped
  to $\ll\Omega\rr_\I = \Omega$ by uniqueness of $\ll\blank\rr$.
\end{proof}

First we define the term signature algebra using which we obtain the
constructors $\con$ for any IIT signature. Then we will define another
signature algebra which provides the eliminator. Before doing these,
we illustrate the idea of both constructions on natural numbers.
\begin{example}\label{eg:nat_con}
  For natural numbers, we will define the constructors $\con$ as the
  following natural number algebra
  $(\mathsf{Nat},\mathsf{zero},\mathsf{suc})$. We write variable names
  instead of de Bruijn indices for readability.
  \begin{alignat*}{5}
    & \mathsf{Nat} && := \Tm^\I\,(\cdot\rhd N:\U\rhd z:\El\,N\rhd s:N\Ra\El\,N)\,(\El\,N) \\
    & \mathsf{zero} && := z \\
    & \mathsf{suc} && := \lambda t. (s\oldapp t)
  \end{alignat*}
  Natural numbers are simply $\I$-terms of type $\El\,N$ in the
  context which is the syntactic signature for natural numbers. In
  this context, the only way to define a term of type $\El\,N$ is to
  use $z$ and $s$, corresponding to the $\mathsf{zero}$ and
  $\mathsf{suc}$ constructors.

@@

\begin{definition}[Term signature algebra $\A\blank\C_{\blank}$]
  For an $M:\SignAlg$ and $\Omega:\Con^M$, we define ${\A M\C}_\Omega$
  which we call the term signature algebra. It is equipped with
  morphisms $\blank^\A : \SignMor\,({\A M\C}_\Omega)\,\A$ and
  $\blank^M : \SignMor\,({\A M\C}_\Omega)\,M$.

  We define ${\A M\C}_\Omega$ by listing its components $\Con$, $\Ty$,
  $\Sub$, and so on, one per row. Each component has three parts
  denoted by $^\A$, $^M$ and $^\C$. The $^\A$ and $^M$ parts are just
  reusing the corresponding components from $\A$ and $M$,
  respectively, and thus the morphisms $\blank^\A$ and $\blank^M$ are
  defined as the obvious projections. We omit the equality components,
  as they come from UIP or are trivial. We also omit the components
  for terms and substitutions as their $^\C$ parts are uninformative
  equational reasonings.
  \begin{alignat*}{10}
    & \Gamma^\A:\Set && \hspace{0.4em} && \Gamma^M:\Con^M && \hspace{0.4em} && \Gamma^\C: \Sub^M\,\Omega\,\Gamma^M\ra \Gamma^\A \\
    & A^\A:\Gamma^\A\ra\Set && && A^M:\Ty^M\,\Gamma^M && && A^\C:(\nu:\Sub^M\,\Omega\,\Gamma^M)\ra \\
    & && && && && \hspace{1em} \Tm^M\,\Omega\,(A^M[\nu])\ra A^\A\,(\Gamma^\C\,\nu) \\
    & \sigma^\A:\Gamma^\A\ra\Delta^\A && && \sigma^M:\Sub^M\,\Gamma^M\,\Delta^M && && \sigma^\C:\Delta^\C\,(\sigma^M\circ\nu)=\sigma^A\,(\Gamma^\C\,\nu) \\
    & t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma && && t^M:\Tm^M\,\Gamma^M\,A^M && && t^\C : A^\C\,\nu\,(t^M[\nu])= t^A\,(\Gamma^\C\,\nu) \\
    & (A[\sigma])^\A\,\gamma := A^\A\,(\sigma^\A\,\gamma) && && (A[\sigma])^M := A^M[\sigma^M]^M && && (A[\sigma])^\C\,\nu\,t := A^\C\,(\sigma^M\circ\nu)\,t \\
    & \cdot^\A := \top && && \cdot^M := \cdot^M && && \cdot^\C\,\nu := \tt \\
    & (\Gamma\rhd A)^\A := && && (\Gamma\rhd A)^M :=  && && (\Gamma\rhd A)^\C\,\nu := \\
    & \hspace{1em} (\gamma:\Gamma^\A)\times A^\A\,\gamma && && \hspace{1em} \Gamma^M\rhd^M A^M && && \hspace{1em} (\Gamma^\C\,(\pi_1\,\nu), A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
    & \U^\A\,\gamma := \Set && && \U^M := \U^M && && \U^\C\nu\,a := \Tm^M\,\Omega\,(\El^M\,a) \\
    & (\El\,a)^\A\,\gamma := a^\A\,\gamma && && (\El\,a)^M := \El^M\,a^M && && (\El\,a)^\C\,\nu\,t := (a^\C\,\nu) \# t \\
    & (\Pi\,a\,B)^\A\,\gamma := && && (\Pi\,a\,B)^M := && && (\Pi\,a\,B)^\C\,\nu\,t :=  \\
    & \hspace{1em} (\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma,\alpha) && && \hspace{1em} \Pi^M\,a^M\,B^M && && \hspace{1em} \lambda\alpha.B^\C\,\big(\nu, (a^\C\,\nu)\#\alpha\big)\,\big(t\oldapp ((a^\C\,\nu) \#\alpha)\big) \\
    & (\Pim\,T\,B)^\A\,\gamma := && && (\Pim\,T\,B)^M := && && (\Pim\,T\,B)^\C\,\nu\,t := \lambda\alpha.(B\,\alpha)^\C\,\nu\,(t\appm\alpha) \\
    & \hspace{1em} (\alpha:T)\ra(B\,\alpha)^\A\,\gamma && && \hspace{1em} \Pim^M\,T\,B^M && && 
  \end{alignat*}
\end{definition}
\begin{example}
  Now given a syntactic signature $\Omega:\Con^\I$, we get the
  constructors as an $\Omega$-algebra by
  $(\ll\Omega\rr_{{\A\I\C}_\Omega})^\C\,\id^\I : \ll\Omega\rr^\A$. If
  $\Omega$ is the syntactic signature for natural numbers, we get the
  constructors as in \Cref{eg:nat_con}.

  An $a : \Tm\,\Omega\,\U$ is a sort term for a signature $\Omega$. If
  $\Omega$ is the syntactic signature for natural numbers, $a$ can
  only be $N$ ($1$ as a de Bruijn index). If $\Omega$ is the syntactic
  signature for $\Con$--$\Ty$ (\Cref{ex:contexts}), $a$ can be
  $Con$, $Ty\oldapp{empty}$,
  $Ty\oldapp({ext}\oldapp {empty}\oldapp(U\oldapp{empty}))$, and so
  on. In any case, for such an $a$, we obtain
  $(\ll a\rr_{{\A\I\C}_\Omega})^\C\,\id^\I : \Tm^\I\,\Omega\,(\El\,a)
  = \ll a\rr^\A\,(\ll\Omega\rr^\C\,\id^\I)$. That is, the algebra
  interpretation of a sort term at the constructors is equal to terms
  of that sort.

  A $t : \Tm\,\Omega\,(\El\,a)$ is a term of a sort type $a$
  constructed using the constructors in $\Omega$. For natural numbers,
  such a $t$ can only be $s$ applied iteratively to $z$. For such a
  $t$, we obtain
  $\ll t\rr^\C\,\id^\I : t = \ll
  t\rr^\A\,(\ll\Omega\rr^\C\,\id^\I)$. That is, a constructor term is
  equal to its algebra interpretation at the constructors. This is
  exactly the equation needed at the end of \Cref{eg:nat_con}.
\end{example}
\begin{remark}
  Given a syntactic signature $\Omega : \Con^\I$, we can use the term
  signature algebra with choices other than $M = \I$. For example, if
  $M = \A$, we usual get the Church encoding by
  $(\ll\Omega\rr_{{\A\A\C}_{\ll\Omega\rr^\A}})^\C\,\id^\A$. In the
  case of natural numbers, this computes (up to curry-uncurry) to
  $(N:\Set)(z:N)(s:N\ra N)\ra N$. Using $M = {\A\M}$ (the signature
  algebra $\A$ extended with the binary version of $\D$ also called
  logical relations, see \cite[Section
  5]{Kaposi:2019:CQI:3302515.3290315}), we obtain the Church encoding
  of Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}.
\end{remark}

\begin{definition}[Eliminator signature algebra $\IE_{\blank}$]
  Given an $\Omega:\Con^\I$,
  $\omega^D : \ll\Omega\rr^\D\,(\ll\Omega\rr^\C\,\id^\I)$, we define a
  the signature algebra $\IE_{\omega^\D}$ as follows. It is equipped
  with a morphism $\blank^\I : \SignMor\,\IE_{\omega^D}\,\I$.

  We define $\IE_{\omega^D}$ by listing its components $\Con$, $\Ty$,
  $\Sub$, and so on, one per row. Each component has two parts denoted
  by $^\I$, and $^\E$. The $^\I$ part just reuses the corresponding
  components of $\I$, thus the morphism $\blank^\I$ is defined as the
  obvious projection. We omit the equality components, as they come
  from UIP or are trivial. We also omit the components for terms and
  substitutions as their $^\E$ parts are uninformative equational
  reasonings.

xxx
\end{definition}

>>>>

We will use Agda-like type theoretic syntax to work in the internal
language of models of ETT:
\begin{itemize}
\item Universes are written $\Set_i$. We usually omit level indices in this paper.
\item $\Pi$ types are notated as $(x : A)\ra B$, or as $A \ra B$ when
  non-dependent. We sometimes omit function arguments, by implicitly
  generalising over variables.
\item $\Sigma$-types, notated either as $(x : A)\times B$, or as
  $\sum\limits_{x} B$ when we want to leave the type of the first
  projection implicit. Projections are either named or given by
  $\proj_1$ and $\proj_2$. We use $A \times B$ for non-dependent
  pairs.
\item We will define indexed W-types \cite{indexedcont} by listing
  their constructors and using pattern matching to eliminate from
  them.
\item The equality (identity) type is written $t = u$, it has a
  constructor $\refl : t = t$, and equality reflection, hence we use
  the same $=$ sign for definitional equality. We occasionally
  indicate by $_{e_1,\dots,e_n \#}t$ that $t$ is well-typed thanks to the
  equalities $e_1$,\dots,$e_n$. We also have uniqueness of identity
  proofs (UIP), expressing $(e : t = t)\ra e = \refl$. Note that
  function extensionality, expressing
  $((x : A) \ra f\,x = g\,x) \ra f = g$ is derivable.
\end{itemize}

@@

    & (t\oldapp u)^\A\,\gamma := t^\A\,\gamma\,(u^\A\,\gamma) && &&  (t\oldapp u)^\D\,\gamma^D := t^\D\,\gamma^D\,(u^\D\,\gamma^D) && &&  (t\oldapp u)^\S\,\gamma^S := _{u^\S\,\gamma^S \#} t^\S\,\gamma^\S\,(u^\A\,\gamma) \\

@@

One may wonder what is the relationship between the two notion of
signatures.
\begin{lemma}
  In a model of ETT supporting the theory of ITT signatures,
  signatures and syntactic signatures are isomorphic.
\end{lemma}
\begin{proof}
  We can turn a $({sig},{sigc}) : \Sign$ into $\Con^\I$ by ${sig}\,\I$
  and an $\Omega:\Con^\I$ into a $\Sign$ by
  $\Big(\lambda M.\ll\Omega\rr_M, \lambda M\,N\,f.\big(f\,\ll\Omega\rr_M =
  (f\circ\ll\blank\rr_M)\,\Omega = \ll\Omega\rr_N\big)\Big)$ where the equality
  proof in the second component comes from uniqueness of the recursor
  (we have to define composition of morphisms $\circ$ for this). The
  compositions of these two maps are the identities: $({sig},{sigc})$
  is mapped to
  $(\lambda M.\ll{sig}\,I\rr_M,\dots) = (\lambda
  M.\ll\blank\rr_M\,({sig}\,I),\dots)$ which is equal to
  $(\lambda M.{sig}\,M,\dots)$ because of ${sigc}$; $\Omega$ is mapped
  to $\ll\Omega\rr_\I = \Omega$ by uniqueness of $\ll\blank\rr$.
\end{proof}

We will define the term signature algebra using which we obtain the
constructors $\con$ for any IIT signature. Then we will define another
signature algebra which provides the eliminator. Before doing these,
we illustrate the idea of both constructions on natural numbers.
\begin{example}\label{eg:nat_con}
  For natural numbers, we will define the constructors $\con$ as the
  following natural number algebra
  $(\mathsf{Nat},\mathsf{zero},\mathsf{suc})$. We write variable names
  instead of de Bruijn indices for readability.
  \begin{alignat*}{5}
    & \mathsf{Nat} && := \Tm^\I\,(\cdot\rhd N:\U\rhd z:\El\,N\rhd s:N\Ra\El\,N)\,(\El\,N) \\
    & \mathsf{zero} && := z \\
    & \mathsf{suc} && := \lambda t. (s\oldapp t)
  \end{alignat*}
  Natural numbers are simply $\I$-terms of type $\El\,N$ in the
  context which is the syntactic signature for natural numbers. In
  this context, the only way to define a term of type $\El\,N$ is to
  use $z$ and $s$, corresponding to the $\mathsf{zero}$ and
  $\mathsf{suc}$ constructors.

@@

\begin{definition}[Term signature algebra $\IC_{\blank}$]
  For an $\Omega:\Con^I$, we define $\IC_\Omega:\SignAlg$ which we
  call the term signature algebra. It is equipped with a morphism
  $\blank^\I : \SignMor\,(\IC_\Omega)\,\I$. We define $\IC_{\Omega}$
  by listing its components $\Con$, $\Ty$, $\Sub$, and so on, one per
  row. Each component has two parts denoted by $^\I$ and $^\C$. The
  $^\I$ part just reuses the corresponding components from $\I$, and
  thus the morphism $\blank^\I$ is defined as the obvious
  projection. We omit the equality components, as they come from UIP
  or are trivial. We also omit the components for terms and
  substitutions as their $^\C$ parts are uninformative equational
  reasonings.
  \begin{alignat*}{10}
    & \Gamma^\I:\Con^\I && \hspace{1em} && \Gamma^\C: \Sub^\I\,\Omega\,\Gamma^\I\ra \ll\Gamma\rr_\A \\
    & A^\I:\Ty^\I\,\Gamma^\I && && A^\C:(\nu:\Sub^\I\,\Omega\,\Gamma^\I)\ra\Tm^\I\,\Omega\,(A^\I[\nu])\ra \ll A\rr_\A\,(\Gamma^\C\,\nu) \\
    & \sigma^\I:\Sub^\I\,\Gamma^\I\,\Delta^\I && && \sigma^\C:\Delta^\C\,(\sigma^\I\circ\nu)=\ll\sigma\rr_\A\,(\Gamma^\C\,\nu) \\
    & t^\I:\Tm^\I\,\Gamma^\I\,A^\I && && t^\C : A^\C\,\nu\,(t^\I[\nu])= \ll t\rr_\A\,(\Gamma^\C\,\nu) \\
    & (A[\sigma])^\I := A^\I[\sigma^\I]^\I && && (A[\sigma])^\C\,\nu\,t := A^\C\,(\sigma^\I\circ\nu)\,t \\
    & \cdot^\I := \cdot^\I && && \cdot^\C\,\nu := \tt \\
    & (\Gamma\rhd A)^\I := \Gamma^\I\rhd^\I A^\I && && (\Gamma\rhd A)^\C\,\nu := (\Gamma^\C\,(\pi_1\,\nu), A^\C\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
    & \U^\I := \U^\I && && \U^\C\nu\,a := \Tm^\I\,\Omega\,(\El^\I\,a) \\
    & (\El\,a)^\I := \El^\I\,a^\I && && (\El\,a)^\C\,\nu\,t := _{a^\C\,\nu\#}t \\
    & (\Pi\,a\,B)^\I := \Pi^\I\,a^\I\,B^\I && && (\Pi\,a\,B)^\C\,\nu\,t := \lambda\alpha.B^\C\,(\nu, _{a^\C\,\nu\#}\alpha)\,(t\oldapp {_{a^\C\,\nu\#}\alpha}) \\
    & (\Pim\,T\,B)^\I := \Pim^\I\,T\,B^\I && && (\Pim\,T\,B)^\C\,\nu\,t := \lambda\alpha.(B\,\alpha)^\C\,\nu\,(t\appm\alpha)
  \end{alignat*}
\end{definition}
\begin{example}\label{ex:constructors}
  Now given a syntactic signature $\Omega:\Con^\I$, we get the
  constructors as an $\Omega$-algebra by
  $\omega:=(\ll\Omega\rr_{\IC_\Omega})^\C\,\id^\I :
  \ll\Omega\rr_\A$. If $\Omega$ is the syntactic signature for natural
  numbers, we get the constructors as in \Cref{eg:nat_con}.

  An $a : \Tm^\I\,\Omega\,\U$ is a sort term for the syntactic
  signature $\Omega$. If $\Omega$ is the syntactic signature for
  natural numbers, $a$ can only be $N$ ($1$ as a de Bruijn index). If
  $\Omega$ is the syntactic signature for $\Con$--$\Ty$
  (\Cref{ex:contexts}), $a$ can be $Con$, $Ty\oldapp{empty}$,
  $Ty\oldapp({ext}\oldapp {empty}\oldapp(U\oldapp{empty}))$, and so
  on. In any case, for such an $a$, we obtain
  $(\ll a\rr_{\IC_\Omega})^\C\,\id^\I : \Tm^\I\,\Omega\,(\El\,a) = \ll
  a\rr_\A\,\omega$. That is, the algebra interpretation of a sort term
  at the constructors is equal to terms of that sort.

  A $t : \Tm^\I\,\Omega\,(\El\,a)$ is a term of a sort type $a$
  constructed using the constructors in $\Omega$. For natural numbers,
  such a $t$ can only be $s$ applied iteratively to $z$. For such a
  $t$, we obtain
  $(\ll t\rr_{\IC_\Omega})^\C\,\id^\I : (t = \ll
  t\rr_\A\,\omega)$. That is, a constructor term is equal to its
  algebra interpretation at the constructors. This is exactly the
  equation needed at the end of \Cref{eg:nat_con}.
\end{example}

\begin{definition}[Eliminator signature algebra $\IE_{\blank}$]
  Given an $\Omega:\Con^\I$, we use the abbreviation
  $\omega := \ll\Omega\rr_{\IC_\Omega}\,\id^\I$ as in
  \Cref{ex:constructors}. Assuming an
  $\omega^D : (\ll\Omega\rr_{\ADS})^\D\,\omega$, we define the
  signature algebra $\IE_{\omega^\D}$. It is equipped with a morphism
  $\blank^\I : \SignMor\,\IE_{\omega^D}\,\I$. We define
  $\IE_{\omega^D}$ by listing its components $\Con$, $\Ty$, $\Sub$,
  and so on, one per row. Each component has two parts denoted by
  $^\I$ and $^\E$. The $^\I$ part just reuses the corresponding
  components of $\I$, thus the morphism $\blank^\I$ is defined as the
  obvious projection. We omit the equality components, as they come
  from UIP or are trivial. We also omit the components for terms and
  substitutions as their $^\E$ parts are uninformative equational
  reasonings.
  \begin{alignat*}{10}
    & \Gamma^\I:\Con^\I && \hspace{1em} && \Gamma^\E:(\nu:\Sub^\I\,\Omega\,\Gamma^\I)\ra \ll\Gamma\rr^\S\,(\ll\nu\rr^\A\,\omega)\,(\ll\nu\rr^\D\,\omega^D) \\
    & A^\I:\Ty^\I\,\Gamma^\I && && A^\E:(\nu:\Sub^\I\,\Omega\,\Gamma^\I)(t:\Tm^\I\,\Omega\,(A^\I[\nu]))\ra \\
    & && && \hspace{1em} \ll A\rr^\S\,(\Gamma^\E\,\nu)\,(\ll t\rr^\A\,\omega)\,(\ll t\rr^\D\,\omega^D) \\
    & \sigma^\I:\Sub^\I\,\Gamma^\I\,\Delta^\I && && \sigma^\E:\Delta^\E\,(\sigma^\I\circ\nu)=\ll\sigma\rr^\S\,(\Gamma^\E\,\nu) \\
    & t^\I:\Tm^\I\,\Gamma^\I\,A^\I && && t^\E : A^\E\,\nu\,(t^\I[\nu])= \ll t\rr^\S\,(\Gamma^\E\,\nu) \\
    & (A[\sigma])^\I := A^\I[\sigma^\I]^\I && && (A[\sigma])^\E\,\nu\,t := A^\E\,(\sigma^\I\circ\nu)\,t \\
    & \cdot^\I := \cdot^\I && && \cdot^\E\,\nu := \tt \\
    & (\Gamma\rhd A)^\I := \Gamma^\I\rhd^\I A^\I && && (\Gamma\rhd A)^\E\,\nu := (\Gamma^\E\,(\pi_1\,\nu), A^\E\,(\pi_1\,\nu)\,(\pi_2\,\nu)) \\
    & \U^\I := \U^\I && && \U^\E\nu\,a := \lambda\alpha . _{\ll\alpha\rr^\C\,\id\#}\big(\ll_{\ll\alpha\rr^\C\,\id\#}\alpha\rr^\D\,\omega^D\big) \\
    & (\El\,a)^\I := \El^\I\,a^\I && && (\El\,a)^\E\,\nu\,t := \big(\ll a\rr^\S\,(\Gamma^\E\,\nu)\,(\ll t\rr^\A\,\omega) \overset{\ll t\rr^\C\,\id}{=} \ll a\rr^\S\,(\Gamma^\E\,\nu)\,t \overset{a^\E\,\nu}{=} \ll t\rr^\D\,\omega^D\big) \\
    & (\Pi\,a\,B)^\I := \Pi^\I\,a^\I\,B^\I && && (\Pi\,a\,B)^\E\,\nu\,t := \\
    & && && \hspace{1em} \lambda\alpha.{_{\ll\alpha\rr^\C\,\id\#}}\big(B^\E\,(\nu,_{\ll a\rr^\C\,\id,\ll\nu\rr^\C\,\id\#}\alpha)\,(t\oldapp {_{\ll a\rr^\C\,\id,\ll\nu\rr^\C\,\id\#}u})\big) \\
    & (\Pim\,T\,B)^\I := \Pim^\I\,T\,B^\I && && (\Pim\,T\,B)^\E\,\nu\,t := \lambda\alpha.(B\,\alpha)^\E\,\nu\,(t\appm\alpha)
  \end{alignat*}
\end{definition}
\begin{example}
  Given the assumptions $\Omega$, $\omega^D$ of $\IE$, we obtain the
  eliminator by
  $\ll\Omega\rr_{\IE_{\omega^\D}}\,\id^\I :
  \ll\Omega\rr^\S\,\omega\,\omega^D$. The eliminator is a section of
  the displayed algebra $\omega^D$, that is, a dependent function
  together with equalities witnessing that all the operations are
  preserved. If $\Omega$ is the syntactic signature for natural
  numbers, we get the eliminator of \Cref{eg:nat_con}.

  For a sort term $a:\Tm^\I\,\Omega\,\U$, the interpretation
  $(\ll a\rr_{\IE_{\omega^D}})^\E\,\id$ says that
  $(\lambda\alpha.\ll\alpha\rr^\D\,\omega^D) = \ll
  a\rr^\S\,(\ll\Omega\rr^\E\,\id)$, that is, the function for the sort
  $a$ in the eliminator section is the displayed algebra
  interpretation at $\omega^D$ (motives and methods). For natural
  numbers, this is the same as
  $\big(\lambda n.\ll n\rr^D\,(N^D,z^D,s^D)\big) = \big(\lambda
  n.\mathsf{elimNat}\,(N^D,z^D,s^D)\,n)\big)$.

  The interpretation of a constructor term
  $t:\Tm^\I\,\Omega\,(\El\,a)$ is uninteresting as it provides an
  equality between two different equality proofs of the computation
  ($\beta$) rule for $t$.
\end{example}

xxx

### 2020-05-03 paper/universaliit.tex
stuff

Let us mention another possible solution for avoiding recursion-recursion:
defining $A\sim\model{A}$ so that it is equivalent to $(e:\rec\,\Gamma
= \model{\Gamma})\times (\rec\,A = e\#\model{A})$.
% whereas the right hand side is of type $\model\Ty\model{\Gamma}$.
% For a substitution $\sigma$ between contexts $\Gamma$ and $\Delta$,
% a first idea consists in defining a relation $\sigma \sim_{\blank} \model{\sigma} : \Gamma
% \sim \model\Gamma \to \Set $ where $\model{A}$ is a semantic type in the
% semantic context $\model{\Gamma}$. We define the relation on types under the hypothesis
% that the contexts are already related.
% it is natural to define the relation with a semantic type $\model{A}$ in a
% semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma = \model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality is
% required for the second one to be well-typed).
% The discussion is similar for terms.
% Then, the ultimate goal is to prove that
% $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
% $\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
% are contractible, i.e., have a unique inhabitant.
%The first step consists in showing that these types are propositions.
In comparison, our approach
% Actually, we take a different approach for the relation on types
% (and terms) which
yields a more concise definition of the relation.
For example, in the case of the universe, this would lead
to the definition
  $  \Uw\,\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
  instead of
  our definition
  $  \Uw\,\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
% Recall that our adapted goal, for types, is to prove that
% $\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
% is inhabited if $\Gamma \sim \model{\Gamma}$ is.
%  This means that in the definition of the relation, we assume
% that contexts are already related, so that we don't need to enforce them to be,
% as it is the case in the original approach.
% We give the two versions of the universe
% case below, to explain the differences.

@@

 := (e_E : \model{E} = \model{\boldsymbol{\cdot}})\times (\model{\delta} = e_E \#

@@

    (\delta = e_E \# \model\sigma \model{\cons} \model{t} )

@@

     (e_B : \model{B} = e_\Delta \# \model{\wk}\,\model{A}) \times

@@

     (\model{t} = e_\Delta,e_B \# \model{\vz})

@@

     (e_C : \model{C} = e_\Delta \# \model{\wk}\,\model{A}) \times

@@

     (\model{t} = e_\Delta,e_C \# \model{\vs} \, \model{n})

@@

      (\model{x} = e_C \# \model{t} \model{\oldapp}\model{u})

@@

      (\model{x} = e_B \# \model{t}\model{\hat{\oldapp}}u)

>>>>

Let us mention another possible solution for avoiding recursion-recursion:
defining $A\sim\model{A}$ so that it is equivalent to $(e:\rec\,\Gamma
= \model{\Gamma})\times (\rec\,A = _{e\#}\model{A})$.
% whereas the right hand side is of type $\model\Ty\model{\Gamma}$.
% For a substitution $\sigma$ between contexts $\Gamma$ and $\Delta$,
% a first idea consists in defining a relation $\sigma \sim_{\blank} \model{\sigma} : \Gamma
% \sim \model\Gamma \to \Set $ where $\model{A}$ is a semantic type in the
% semantic context $\model{\Gamma}$. We define the relation on types under the hypothesis
% that the contexts are already related.
% it is natural to define the relation with a semantic type $\model{A}$ in a
% semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma = \model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality is
% required for the second one to be well-typed).
% The discussion is similar for terms.
% Then, the ultimate goal is to prove that
% $\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
% $\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
% are contractible, i.e., have a unique inhabitant.
%The first step consists in showing that these types are propositions.
In comparison, our approach
% Actually, we take a different approach for the relation on types
% (and terms) which
yields a more concise definition of the relation.
For example, in the case of the universe, this would lead
to the definition
  $  \Uw\,\wf{\Gamma} \sim \model{A} := (\wf{\Gamma}\sim\model{\Gamma}) \times (\model{A} = \model{\U})$,
  instead of
  our definition
  $  \Uw\,\wf{\Gamma} \sim \model{A} := (\model{A} = \model{\U})$.
% Recall that our adapted goal, for types, is to prove that
% $\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
% is inhabited if $\Gamma \sim \model{\Gamma}$ is.
%  This means that in the definition of the relation, we assume
% that contexts are already related, so that we don't need to enforce them to be,
% as it is the case in the original approach.
% We give the two versions of the universe
% case below, to explain the differences.

@@

 := (e_E : \model{E} = \model{\boldsymbol{\cdot}})\times (\model{\delta} = _{e_E \#}

@@

    (\delta = _{e_E \#} \model\sigma \model{\cons} \model{t} )

@@

     (e_B : \model{B} = _{e_\Delta \#} \model{\wk}\,\model{A}) \times

@@

     (\model{t} = _{e_\Delta,e_B \#} \model{\vz})

@@

     (e_C : \model{C} = _{e_\Delta \#} \model{\wk}\,\model{A}) \times

@@

     (\model{t} = _{e_\Delta,e_C \#} \model{\vs} \, \model{n})

@@

      (\model{x} = _{e_C \#} \model{t} \model{\oldapp}\model{u})

@@

      (\model{x} = _{e_B \#} \model{t}\model{\hat{\oldapp}}u)
